### 2018-03-16 submission/manuscript.Rmd
fix typos in intro

This study will specifically address how cycle number and high fidelity (HiFi) DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing. First, it is important to differentiate between bias, reproducibility, and standardization since often times these three can be confused and used interchangeably with each other. Bias can change the observed results in a way that is reproducible and standardized. For example, if one group uses one brand of DNA extraction kit for their 16S rRNA gene sequencing, their results may be biased versus another group not using the same brand kit but within their group they can still have reproducible results. Therefore, standardization of 16S rRNA gene sequencing methods to increase reproducibility can still be problematic due to bias. Determining how different aspects of a 16S rRNA gene sequencing workflow could bias the observed results is critical for the interpretation of specific studies in the borader context of the overall field.

A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. Generally, not using a preservation media and leaving samples at room temperature has been shown to cause overgrowth of low abundance members of the fecal bacterial community [@Amir2017]. Similarily, this overgrowth could still occur if the preservation media does not adequately inhibit growth [@preservation_Luo_2016; @preservation_Song_2016]. Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [@Gorzelak2015]. Additionally, reagent contamination has been shown to add community members and the contribution of these contaminant members grows larger with lower biomass samples [@contamination_Salter2014]. Recent studies have shown that the majority of these biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [@preservation_Song_2016; @storage_Bassis_2017]. However, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [@Eckert1991; @Burkardt2000]. Many of these sources of biases are made worse as cycle number increases [@Wang1996; @Haas2011; @Kebschull2015]. For example, the selective amplification of AT-rich over GC-rich sequences can increase the difference between these types of 16S rRNA genes and over-emphasizing the actual difference [@polz_bias_1998]. Both amplification error and non-specific amplification (e.g. incorrect amplicon size products) can also increase as cycle number increases which can drastically change commonly used  diversity measures [@BautistadelosSantos2016; @Acinas2005]. Additionally, chimeras can form from an aborted extension followed by subsequent priming error and secondary extension and even if it has not been explicitly shown to increase with cycle number it is easy to postulate that this does happen and will also artificially increase community diversity [@Haas2011].

There are also intrinsic properties to primers and DNA polymerases chosen that can introduce bias. Primers have variable region dependent binding affinities for different bacteria and depending on the primer pair do not detect specific bacteria (e.g. V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)]. Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capcity [@polymerase_Ishino_2014]. Interestingly, the influence that these different DNA polymerases can have on the observed results have not been well studied like some of the other previously mentioned sources of PCR-based bias.

A recent study found clear differences between normal and high fidelity (HiFi) DNA polymerase and that optimization of the PCR protocol could reduce error and chimera generation [@taq_Gohl_2016]. This study also found that regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline [@taq_Gohl_2016]. It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on HiFi DNA polymerases. There is some reason to think that this may be the case since many of these HiFi DNA polymerases come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study, we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

>>>>

This study will specifically address how cycle number and high fidelity (HiFi) DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing. First, it is important to differentiate between bias, reproducibility, and standardization since often times these three can be confused and used interchangeably with each other. Bias can change the observed results in a way that is reproducible and standardized. For example, if one group uses one brand of DNA extraction kit for their 16S rRNA gene sequencing, their results may be biased versus another group not using the same brand kit but within their group they can still have reproducible results. Therefore, standardization of 16S rRNA gene sequencing methods to increase reproducibility can still be problematic due to bias. Determining how different aspects of a 16S rRNA gene sequencing workflow could bias the observed results is critical for the interpretation of specific studies in the broader context of the overall field.

A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. Generally, not using a preservation media and leaving samples at room temperature has been shown to cause overgrowth of low abundance members of the fecal bacterial community [@Amir2017]. Similarily, this overgrowth can still occur if the preservation media does not adequately inhibit growth [@preservation_Luo_2016; @preservation_Song_2016]. Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [@Gorzelak2015]. Additionally, reagent contamination has been shown to add community members and the contribution of these contaminant members grows larger with lower biomass samples [@contamination_Salter2014]. Recent studies have shown that the majority of these biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [@preservation_Song_2016; @storage_Bassis_2017]. However, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [@Eckert1991; @Burkardt2000]. Many of these sources of biases are made worse as cycle number increases [@Wang1996; @Haas2011; @Kebschull2015]. For example, the selective amplification of AT-rich over GC-rich sequences can exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC [@polz_bias_1998]. Both amplification error and non-specific amplification (e.g. incorrect amplicon size products) can also increase as cycle number increases which can drastically change commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005]. Additionally, chimeras can form from an aborted extension step followed by a subsequent priming error and secondary extension and will also artificially increase community diversity [@Haas2011].

There are also intrinsic properties to primers and DNA polymerases chosen that can introduce bias. Primers have variable region dependent binding affinities for different bacteria and depending on the primer pair do not detect specific bacteria (e.g. V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)]. Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014]. Interestingly, the influence that these different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other previously mentioned sources of PCR-based bias.

A recent study found clear differences between normal and high fidelity (HiFi) DNA polymerase and that optimization of the PCR protocol could reduce error and chimera generation [@taq_Gohl_2016]. This study also found that regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline [@taq_Gohl_2016]. It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on the type of HiFi DNA polymerase. There is some reason to think that this may be the case since many of these HiFi DNA polymerases come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study, we examine if any of five different types of HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

### 2018-03-16 submission/manuscript.tex
fix typos in intro

This study will specifically address how cycle number and high fidelity
(HiFi) DNA polymerases can bias observed bacterial community results
derived from 16S rRNA gene sequencing. First, it is important to
differentiate between bias, reproducibility, and standardization since
often times these three can be confused and used interchangeably with
each other. Bias can change the observed results in a way that is
reproducible and standardized. For example, if one group uses one brand
of DNA extraction kit for their 16S rRNA gene sequencing, their results
may be biased versus another group not using the same brand kit but
within their group they can still have reproducible results. Therefore,
standardization of 16S rRNA gene sequencing methods to increase
reproducibility can still be problematic due to bias. Determining how
different aspects of a 16S rRNA gene sequencing workflow could bias the
observed results is critical for the interpretation of specific studies
in the borader context of the overall field.

A typical 16S rRNA gene sequencing workflow can be divided into
preservation, extraction, PCR, and sequencing steps. Generally, not
using a preservation media and leaving samples at room temperature has
been shown to cause overgrowth of low abundance members of the fecal
bacterial community (Amir et al., 2017). Similarily, this overgrowth
could still occur if the preservation media does not adequately inhibit
growth (Song et al., 2016; Luo et al., 2016). Reports have also shown
that changes in specific community members might occur due to differing
susceptibility to freeze thaw cycles amongst microbes (Gorzelak et al.,
2015). Additionally, reagent contamination has been shown to add
community members and the contribution of these contaminant members
grows larger with lower biomass samples (Salter et al., 2014). Recent
studies have shown that the majority of these biases due to either
preservation or extraction tend to be smaller than the overall
biological signal being measured (Song et al., 2016; Bassis et al.,
2017). However, the contribution of PCR bias to this overall workflow is
not well characterized since these studies use the same PCR approach
while varying preservation or extraction method.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
important because a large body of literature shows that there are a
variety of steps during PCR that can change the observed results (Eckert
\& Kunkel, 1991; Burkardt, 2000). Many of these sources of biases are
made worse as cycle number increases (Wang \& Wang, 1996; Haas et al.,
2011; Kebschull \& Zador, 2015). For example, the selective
amplification of AT-rich over GC-rich sequences can increase the
difference between these types of 16S rRNA genes and over-emphasizing
the actual difference (Polz \& Cavanaugh, 1998). Both amplification
error and non-specific amplification (e.g.~incorrect amplicon size
products) can also increase as cycle number increases which can
drastically change commonly used diversity measures (Acinas et al.,
2005; Santos et al., 2016). Additionally, chimeras can form from an
aborted extension followed by subsequent priming error and secondary
extension and even if it has not been explicitly shown to increase with
cycle number it is easy to postulate that this does happen and will also
artificially increase community diversity (Haas et al., 2011).

There are also intrinsic properties to primers and DNA polymerases
chosen that can introduce bias. Primers have variable region dependent
binding affinities for different bacteria and depending on the primer
pair do not detect specific bacteria (e.g.~V1-V3 does not detect
\emph{Haemophilus influenzae} and V3-V5 does not detect
\emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
al., 2016). Additionally, there are multiple families of DNA polymerases
that have their own error rate and proof reading capcity (Ishino \&
Ishino, 2014). Interestingly, the influence that these different DNA
polymerases can have on the observed results have not been well studied
like some of the other previously mentioned sources of PCR-based bias.

A recent study found clear differences between normal and high fidelity
(HiFi) DNA polymerase and that optimization of the PCR protocol could
reduce error and chimera generation (Gohl et al., 2016). This study also
found that regardless of DNA polymerase, the number of Operational
Taxonomic Units (OTUs) or taxa generated were not easily reduced using
the authors chosen bioinformatic pipeline (Gohl et al., 2016). It is
natural to extend this line of inquiry and ask if biases in the number
of OTUs and chimeras are also dependent on HiFi DNA polymerases. There
is some reason to think that this may be the case since many of these
HiFi DNA polymerases come from different families (e.g. \emph{Taq}
belongs to the family A polymerases) and may intrinsically have
different error rates that cannot be completely removed with
modifications (Ishino \& Ishino, 2014). In this study, we critically
examine if any of five different HiFi DNA polymerases introduce
significant biases into 16S rRNA gene surveys, if this is a cycle
dependent phenomenon, and whether they can be removed using a standard
bioinformatic pipeline.

>>>>

This study will specifically address how cycle number and high fidelity
(HiFi) DNA polymerases can bias observed bacterial community results
derived from 16S rRNA gene sequencing. First, it is important to
differentiate between bias, reproducibility, and standardization since
often times these three can be confused and used interchangeably with
each other. Bias can change the observed results in a way that is
reproducible and standardized. For example, if one group uses one brand
of DNA extraction kit for their 16S rRNA gene sequencing, their results
may be biased versus another group not using the same brand kit but
within their group they can still have reproducible results. Therefore,
standardization of 16S rRNA gene sequencing methods to increase
reproducibility can still be problematic due to bias. Determining how
different aspects of a 16S rRNA gene sequencing workflow could bias the
observed results is critical for the interpretation of specific studies
in the broader context of the overall field.

A typical 16S rRNA gene sequencing workflow can be divided into
preservation, extraction, PCR, and sequencing steps. Generally, not
using a preservation media and leaving samples at room temperature has
been shown to cause overgrowth of low abundance members of the fecal
bacterial community (Amir et al., 2017). Similarily, this overgrowth can
still occur if the preservation media does not adequately inhibit growth
(Song et al., 2016; Luo et al., 2016). Reports have also shown that
changes in specific community members might occur due to differing
susceptibility to freeze thaw cycles amongst microbes (Gorzelak et al.,
2015). Additionally, reagent contamination has been shown to add
community members and the contribution of these contaminant members
grows larger with lower biomass samples (Salter et al., 2014). Recent
studies have shown that the majority of these biases due to either
preservation or extraction tend to be smaller than the overall
biological signal being measured (Song et al., 2016; Bassis et al.,
2017). However, the contribution of PCR bias to this overall workflow is
not well characterized since these studies use the same PCR approach
while varying preservation or extraction method.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
important because a large body of literature shows that there are a
variety of steps during PCR that can change the observed results (Eckert
\& Kunkel, 1991; Burkardt, 2000). Many of these sources of biases are
made worse as cycle number increases (Wang \& Wang, 1996; Haas et al.,
2011; Kebschull \& Zador, 2015). For example, the selective
amplification of AT-rich over GC-rich sequences can exaggerate the
difference between 16S rRNA genes higher in AT versus those higher in GC
(Polz \& Cavanaugh, 1998). Both amplification error and non-specific
amplification (e.g.~incorrect amplicon size products) can also increase
as cycle number increases which can drastically change commonly used
diversity measures (Acinas et al., 2005; Santos et al., 2016).
Additionally, chimeras can form from an aborted extension step followed
by a subsequent priming error and secondary extension and will also
artificially increase community diversity (Haas et al., 2011).

There are also intrinsic properties to primers and DNA polymerases
chosen that can introduce bias. Primers have variable region dependent
binding affinities for different bacteria and depending on the primer
pair do not detect specific bacteria (e.g.~V1-V3 does not detect
\emph{Haemophilus influenzae} and V3-V5 does not detect
\emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
al., 2016). Additionally, there are multiple families of DNA polymerases
that have their own error rate and proof reading capacity (Ishino \&
Ishino, 2014). Interestingly, the influence that these different DNA
polymerases can have on the observed 16S rRNA gene sequencing results
have not been well studied like some of the other previously mentioned
sources of PCR-based bias.

A recent study found clear differences between normal and high fidelity
(HiFi) DNA polymerase and that optimization of the PCR protocol could
reduce error and chimera generation (Gohl et al., 2016). This study also
found that regardless of DNA polymerase, the number of Operational
Taxonomic Units (OTUs) or taxa generated were not easily reduced using
the authors chosen bioinformatic pipeline (Gohl et al., 2016). It is
natural to extend this line of inquiry and ask if biases in the number
of OTUs and chimeras are also dependent on the type of HiFi DNA
polymerase. There is some reason to think that this may be the case
since many of these HiFi DNA polymerases come from different families
(e.g. \emph{Taq} belongs to the family A polymerases) and may
intrinsically have different error rates that cannot be completely
removed with modifications (Ishino \& Ishino, 2014). In this study, we
examine if any of five different types of HiFi DNA polymerases introduce
significant biases into 16S rRNA gene surveys, if this is a cycle
dependent phenomenon, and whether they can be removed using a standard
bioinformatic pipeline.
